Secure Composition of Robust and Optimising Compilers

### 2025-02-13 icfp/sections/bg/hsat.tex
Sketch out hsat

A hyperproperty ($\varHProp$) is a set of program behaviours, i.e., a set of sets of traces, which have been referred to as {\em class} (see \Cref{subsec:bg:tprop}).
Much like trace-based properties, hyperproperties can also be classified into different categories, e.g., hypersafety or hyperliveness~\cite{}.

@@

%TODO: this is wrong. rework to explain it via program behaviours instead
\begin{definition}[Hyper-Satisfaction]\label{def:prophsat}
  $\sat{\wholeprogvar}{\varHProp}$
  $\isdef$
  $\forall\varProp\in\varHProp,\sat{\wholeprogvar}{\varProp}$.
\end{definition}
\begin{definition}[Robust Hyper-Satisfaction]\label{def:proprhsat}
  $\rsat{\progvar}{\varHProp}$
  $\isdef$ 
  $\forall\varProp\in\varHProp,\rsat{\progvar}{\varProp}$.
\end{definition}

The definitions in \Cref{subsec:bg:rtp} are trivially extended to the hyperproperty case, so we omit it.
However, it is worth noting that program refinement does not commute with hyperproperty refinement, while it does so for traces.
% TODO: write more
Consider 
\[
\begin{array}{rcl}
  \operatorname{behav}\left(\src{p}\right) &=& \left\{ \src{\event_0} \right\} \\
  \operatorname{behav}\left(\src{p'}\right) &=& \left\{ \src{\event_0}, \src{\event_1} \right\} \\[.33cm]
  \varHProp &=& \left\{ \left\{ \src{\event_0} \right\}, \left\{ \src{\event_1} \right\} \right\} \\
  \varHProp' &=& \left\{ \left\{ \src{\event_1} \right\} \right\} \\
\end{array}
\]
Clearly, $\operatorname{behav}(\src{p})\subseteq\operatorname{behav}(\src{p'})$, so $\src{p}$ {\em refines} $\src{p'}$.
Similarly, $\varHProp'\subseteq\varHProp$.
Note that $\sat{\src{p}}{\varHProp}$, but $\not\sat{\src{p}}{\varHProp'}$.
This does not break for ordinary trace-based properties, since refinement coincides with satisfaction.

Based off of \Cref{def:prophsat,def:propsat}, it is well known that an ordinary trace-based property can be lifted to an equivalent hyperproperty by means of simply taking its powerset~\cite{clarkson2008hyper}.

>>>>

A hyperproperty ($\varHProp$) is a set of program behaviours, i.e., a set of sets of traces, which have been referred to as {\em class} (see \Cref{subsec:bg:tprop}).
Much like trace-based properties, hyperproperties can also be classified into different categories, e.g., hypersafety or hyperliveness~\cite{clarkson2008hyper}.
Based off of \Cref{def:prophsat,def:propsat}, it is well known that an ordinary trace-based property can be lifted to an equivalent hyperproperty by means of simply taking its powerset~\cite{clarkson2008hyper}.

@@

\begin{definition}[Hyper-Satisfaction]\label{def:prophsat}
  $\sat{\wholeprogvar}{\varHProp}$
  $\isdef$
  $\behav{\wholeprogvar}\in\varHProp$.
\end{definition}
\begin{definition}[Robust Hyper-Satisfaction]\label{def:proprhsat}
  $\rsat{\progvar}{\varHProp}$
  $\isdef$ 
  $\forall\contextvar\ \wholeprogvar,\text{if}\ \link{\contextvar}{\progvar}=\wholeprogvar, \text{then}\sat{\wholeprogvar}{\varHProp}$.
\end{definition}

The definitions in \Cref{subsec:bg:rtp} are trivially extended to the hyperproperty case, so we omit them.
However, it is worth noting that program refinement does not commute with hyper-satisfaction, while it does commute for ordinary satisfaction.
To see this, consider the following example:
 %\[
 %\begin{array}{c}
 %  \operatorname{behav}\left(\src{\wholeprogvar}\right) = \left\{ \src{\trace[_0]} \right\} \quad
 %  \operatorname{behav}\left(\src{\wholeprogvar'}\right) = \left\{ \src{\trace[_0]}, \src{\trace[_1]} \right\} \qquad
 %  \varHProp = \left\{ \left\{ \src{\trace[_0]} \right\}, \left\{ \src{\trace[_1]} \right\} \right\} \quad
 %  \varHProp' = \left\{ \left\{ \src{\trace[_1]} \right\} \right\} \\
 %\end{array}
 %\]
\begin{center}
\begin{tikzpicture}
  \node (a) {$\behav{\src{\wholeprogvar}}=\left\{\src{\trace[_0]}\right\}$};
  \node[right=2cm of a] (b) {$\behav{\src{\wholeprogvar[']}}=\left\{\src{\trace[_0]},\src{\trace[_1]}\right\}$};
  \node[below=1cm of b] (d) {$\varHProp=\left\{\left\{\src{\trace[_0]},\src{\trace[_1]}\right\},\left\{\src{\trace[_1]}\right\}\right\}$};

  \node[right=.8cm of a] (ab) {$\subseteq$};
  \node[below=.5cm of b,rotate=270] (bd) {$\in$};
  \node[below=.3cm of ab,rotate=-33] (ad) {$\not\in$};
\end{tikzpicture}
\end{center}

Clearly, $\operatorname{behav}(\src{\wholeprogvar})\subseteq\operatorname{behav}(\src{\wholeprogvar'})$, so $\src{\wholeprogvar}$ {\em refines} $\src{\wholeprogvar'}$.
However, while $\src{\wholeprogvar[']}$ satisfies $\varHProp$ ($\sat{\src{\wholeprogvar[']}}{\varHProp}$), we have that $\not\sat{\src{\wholeprogvar}}{\varHProp}$.
This does not break for ordinary trace-based properties, since refinement coincides with satisfaction, i.e., taking subsets is a transitive operation.

### 2025-02-13 icfp/sections/bg/sat.tex
Sketch out hsat

Running a (whole) program yields a trace ($\trace$), i.e., a potentially infinite sequence\footnote{
Throughout the paper, sequences are indicated with an overbar (i.e., $\trace$), empty sequences with $\hole{\cdot}$, and concatenation of sequences $\trace[_{1}],\trace[_{2}]$ as $\trace[_{1}]\cdot\trace[_{2}]$.
Prepending element $\event$ to a sequence $\trace$ uses the same notation: $\event\cdot\trace$.
} of events ($\event$).
Events include security-relevant actions (e.g., reading from and writing to memory, as detailed in \Cref{sec:compprop}), the unobservable event ($\emptyevent$), and a crash ($\lightning$).
For a given program, we refer to the set of traces as its behaviour. 
Moreover, sets of traces are called properties ($\varProp$) which describe admissible program behaviours. % ascribing what said property considers valid.
% 
The set of all properties can be split-up into different {\em classes} ($\class$), i.e., safety, liveness, and neither safety nor liveness~\cite{clarkson2008hyper}, so a class is simply a set of properties.
% 
%The compositionality framework (\Cref{sec:sequential}) presented in this paper considers arbitrary classes, while the case-study (\Cref{sec:casestud:rtp}) fixes them to concrete instances of safety properties, since it is decidable whether a trace satisfies a safety property with just a finite trace (i.e., a \emph{prefix}).

As an example, consider the trace $\ev{Dealloc\ \loc}\cdot\ev{Read\ \loc\ 1729}\cdot\dots$ which describes the interaction with memory where the deallocation of an address ($\loc$) precedes a read (of some value $1729$) at that address in memory.
% 
This behaviour is insecure w.r.t. the canonical notion of \gls*{tms} dictating no use-after-frees~\cite{nagarakatte2010cets,azevedo2018meaningsofms}, because it reads from a memory location that was freed already.
The first two events in the shown trace are enough to decide that the trace does not satisfy \gls*{tms}, since there is no way to append events to this prefix which would result in the trace being admissible for \gls*{tms}.

@@

\begin{definition}[Property Satisfaction]\label{def:propsat}
  \bul{$\sat{\wholeprogvar}{\varProp}$}
  $\isdef$
  if \iul{$\forall\trace,\tracegen{\wholeprogvar}{\trace}$},
  then \oul{$\trace\in\varProp$}.
\end{definition}

>>>>

Running a (whole) program ($\wholeprogvar$) yields a trace ($\trace$), i.e., a potentially infinite sequence\footnote{
Throughout the paper, sequences are indicated with an overbar (i.e., $\trace$), empty sequences with $\hole{\cdot}$, and concatenation of sequences $\trace[_{1}],\trace[_{2}]$ as $\trace[_{1}]\cdot\trace[_{2}]$.
Prepending element $\event$ to a sequence $\trace$ uses the same notation: $\event\cdot\trace$.
} of events ($\event$).
Events include security-relevant actions (e.g., reading from and writing to memory, as detailed in \Cref{sec:compprop}), the unobservable event ($\emptyevent$), and a crash ($\lightning$).
For a given program, we refer to the set of traces as its behaviour $\behav{\wholeprogvar}$. 
Moreover, sets of traces are called properties ($\varProp$) which describe admissible program behaviours. % ascribing what said property considers valid.
% 
The set of all properties can be split-up into different {\em classes} ($\class$), i.e., safety, liveness, and neither safety nor liveness~\cite{clarkson2008hyper}, so a class is simply a set of properties.
% 
%The compositionality framework (\Cref{sec:sequential}) presented in this paper considers arbitrary classes, while the case-study (\Cref{sec:casestud:rtp}) fixes them to concrete instances of safety properties, since it is decidable whether a trace satisfies a safety property with just a finite trace (i.e., a \emph{prefix}).

As an example for a safety property, consider the trace $\ev{Dealloc\ \loc}\cdot\ev{Read\ \loc\ 1729}\cdot\dots$ which describes the interaction with memory where the deallocation of an address ($\loc$) precedes a read (of some value $1729$) at that address in memory.
% 
This behaviour is insecure w.r.t. the canonical notion of \gls*{tms} dictating no use-after-frees~\cite{nagarakatte2010cets,azevedo2018meaningsofms}, because it reads from a memory location that was freed already.
The first two events in the shown trace are enough to decide that the trace does not satisfy \gls*{tms}, since there is no way to append events to this prefix which would result in the trace being admissible for \gls*{tms}.

@@

\begin{definition}[Property Satisfaction]\label{def:propsat}
  \bul{$\sat{\wholeprogvar}{\varProp}$}
  $\isdef$
  if \iul{$\forall\trace,\tracegen{\wholeprogvar}{\trace}$},
  then \oul{$\trace\in\varProp$} $\equiv\behav{\wholeprogvar}\subseteq\varProp$.
\end{definition}
